#lab_12
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Overview

There is one main goal of this lab:

* To apply your knowledge of basic web scraping.
* To integrate scraped data with other data pulled from API packages.

You will do this by scraping data from the Maryland State Board of Election website, and connecting it with U.S. Census data pulled from the Tidycensus package to answer questions that can be answered with one or more data sets. 

You can find election results here: https://elections.maryland.gov/elections/2022/general_results/

## Load libraries and settings

**Task**: load the tidyverse, tidycensus, rvest and janitor.  Store your Census api key.

```{r}
library(tidyverse)
library(tidycensus)
library(rvest)
library(janitor)
```

```{r}
#census key
census_api_key("423292f0a0d4b5f0fe4ce722f283f6af07c21c84", install=TRUE, overwrite=TRUE)

readRenviron("~/.Renviron")
```



## Questions

**Q1** Which candidate, Dan Cox or Wes Moore, got the lowest percentage of their votes on Election Day? What does that suggest? Write up your answer in a sentence. 
**A1**
```{r}
gov_race_results <- "https://elections.maryland.gov/elections/2022/general_results/gen_results_2022_1.html"

#read in results as html table

gov_race_results <- gov_race_results %>%   
  read_html() %>%   
  html_table() 

#extract dataframe, clean number columns, calculate pct on election day vote

gov_race_results <- gov_race_results[[1]] %>%  
  clean_names() %>%  
  mutate(across(c(3,4,5,7,8), parse_number)) %>%  
  mutate(pct_election_day = round(election_day/total*100,2)) %>%  
  slice(-7) %>%  
  arrange(pct_election_day) %>%  
  select(name, pct_election_day) 

```


**Q2** How many counties did Chris Van Hollen win in Maryland's Senate race? 
**A2** 
```{r}
#load data
sen_race_results <- "https://elections.maryland.gov/elections/2022/general_results/gen_detail_results_2022_4_1.html"

sen_race_results <- sen_race_results %>% 
  read_html() %>% 
  html_table() 

sen_race_results <- sen_race_results[[1]] %>% 
  clean_names() %>%  
  slice(-25) %>% 
  clean_names() %>%
  mutate(across(c(2:5), parse_number)) %>% 
  rename(chaffee = 2, van_hollen = 3) %>%
  select(jurisdiction, chaffee, van_hollen) %>%
  mutate(
    senate_results = case_when(
      van_hollen > chaffee ~ "Van Hollen",
      chaffee > van_hollen ~ "Chaffee"
    )
  )

view(sen_race_results)
#Van Hollen won in Anne Arundel, Baltimore City, Baltimore County, Charles, Frederick, Howard, Kent, Montgomery, Prince George's and Talbot counties.
```

**Q3** What percentage of the vote did Chris Van Hollen win in Maryland's wealthiest county by median income. (You must use the Tidycensus package to answer this question).
**A3** 
```{r}
#get the wealthiest county by median income
acs_vars <- load_variables("acs5", year=2020)
md <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2020) %>%
  arrange(desc(estimate))
#howard county is the wealthiest county in maryland

sen_race_results_county <- sen_race_results %>% 
  mutate(van_hollen_pct = (van_hollen/(chaffee+van_hollen)*100))

sen_race_results_howard <- sen_race_results_county %>% 
  filter(jurisdiction == "Howard")

view(sen_race_results_howard)

#Van Hollen had about 71% of the total votes from Howard County.
```

