How does median household income and education correlate, specifically through graduation rates in Baltimore City and scores in English language exams?

Let's start with some background:
Household median income is essentially the median income of a certain area in a city, state or community. Historically, districts have been redrawn -- in a process known as redlining -- to ensure that some areas had a larger white population than others. While this is now an illegal process, some of its effects may still be felt today -- especially in Baltimore. This data analysis could be used to understand how some effects are still seen, especially with drop out rates, test scores, median income and race.  

Let's load our packages:

```{r}
#load janitor, tidyverse and formattable
library(janitor)
library(tidyverse)
library(formattable)
library(ggplot2)
library(ggthemes)
```
Next, let's load our data. 
```{r}
#clean names
#load baltimore race data and baltimore income data
#load md english scores, grade enrollment, dropout rate and diplomas certificates
baltcity_race <- read_csv("baltcity_race_8_13.csv") %>% 
  clean_names()

baltcity_income <- read_csv("baltcity_income_clean.csv") %>% 
  clean_names()

x2020_dropout_rate <- read_csv("Annual_Dropout_Rate_2020.2.csv") %>% 
  clean_names()

x2020_grade_enrollment <- read_csv("Enrollment_By_Grade_2020.csv") %>% 
  clean_names()

x2020_english_scores <- read_csv("2020_ELPA2_Summary.csv") %>% 
  clean_names()

x2020_diplomas_certificates <- read_csv("2020_Diplomas_and_Certificates.csv") %>% 
  clean_names()
```

What areas have the highest median household income?
```{r}
#get the top 10 neighborhoods with the highest median income in 2020
top10_baltcity_income <- baltcity_income %>% 
  select(neighborhood, x2020, census) %>% 
  slice_max(x2020, n=10)

view(top10_baltcity_income)
```
#North Baltimore/Guilford/Homeland has the highest median household income. 


What about the lowest?
```{r}
low10_baltcity_income <- baltcity_income %>% 
  select(neighborhood, x2020, census) %>% 
  slice_min(x2020, n=10)

view(low10_baltcity_income)
```
#Upton/Druid Heights is the neighborhood with the lowest median household income. 

Let's grab the locations in the city with the top white population as well.
```{r}
#cross reference that with the top 10 places with the most white population
top10_white <- baltcity_race %>% 
  select(census_tract,x2020_white) %>% 
  slice_max(x2020_white, n=10)

view(top10_white)
```
#Census Tract 2712 has the largest white population. 

Also, let's grab the lowest 10 white population.
```{r}
low10_white <- baltcity_race %>% 
  select(census_tract,x2020_white) %>% 
  slice_min(x2020_white, n=10)

view(low10_white)
```
#Census Tract 1502 has the lowest white population.

How does this compare to schools with the highest graduation rates?
```{r}
#Filter x2020_diplomas_certificates for Baltimore City, then check on geoids of the schools with the highest graduation rates
baltcity_grads <- x2020_diplomas_certificates %>% 
  filter(lss_name== "Baltimore City") %>% 
  select(lss_name,school_name, total_hs_completers) %>% 
 slice_max(total_hs_completers, n=10)

view(baltcity_grads)

```
#The Baltimore School for the Arts has the highest count of diplomas/certificates out of schools in Baltimore City. Its census tract is not within the top 10 tracts with the largest white population.

How are English Language scores in Baltimore City? Is there a specific school that has the lowest scores, and in what area of the city is that located? Do students who speak another language get the help they need for high school graduation? 

```{r}
#find lowest English scores by city
x2020_english_scores$proficient_count<- as.numeric(x2020_english_scores$proficient_count)

x2020_baltcity_english_scores <- x2020_english_scores %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, proficient_count) %>% 
  slice_min(proficient_count, n=10)

view(x2020_baltcity_english_scores)
```

Now to find the schools with the highest proficiency count in Baltimore City.
```{r}
x2020_english_scores$proficient_count<- as.numeric(x2020_english_scores$proficient_count)
  
x2020_baltcity_english_scores <- x2020_english_scores %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, proficient_count) %>% 
 slice_max(proficient_count, n=10)

view(x2020_baltcity_english_scores)
```
#John Ruhrah Elementary/Middle has the lowest English scores in Baltimore City. 

What schools in Baltimore City have the highest drop out rate in 2020?
```{r}
x2020_dropout_rate$dropout_rate_numer <- as.numeric(x2020_dropout_rate$dropout_rate_numer)
  
baltcity_dropout <- x2020_dropout_rate %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, dropout_rate_numer) %>% 
  slice_max(dropout_rate_numer, n=10)

view(baltcity_dropout)
```
#Excel Academy has the highest dropout count besides from the count for the entirety of Baltimore City schools. 

Let's make a table with findings of the Baltimore City dropout count:
```{r}
baltcity_dropout %>% 
  ggplot(aes(x=school_name, y=dropout_rate_numer, weight=dropout_rate_numer)) +
  theme_economist() +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
    labs(
    title="Schools by dropout rate Baltimore City",
    x = "School Name",
    y = "Dropout Count",
    caption = "source: Baltimore City data")
  
```

