**Elizabeth Alspach, Final data notebook assignment March 12, 2024**

How does median household income and education correlate, specifically through graduation rates in Baltimore City and scores in English language exams?

Let's start with some background:
Household median income is essentially the median income of a certain area in a city, state or community. Historically, districts have been redrawn -- in a process known as redlining -- to ensure that some areas had a larger white population than others. While this is now an illegal process, some of its effects may still be felt today -- especially in Baltimore. This data analysis could be used to understand how some effects are still seen, especially with drop out rates, test scores, median income and race.  

Let's load our packages:

```{r}
#load janitor, tidyverse and formattable
library(janitor)
library(tidyverse)
library(formattable)
library(ggplot2)
library(ggthemes)
```
Next, let's load our data. 
```{r}
#clean names
#load baltimore race data and baltimore income data
#load md english scores, grade enrollment, dropout rate and diplomas certificates
baltcity_race <- read_csv("baltcity_race_8_13.csv") %>% 
  clean_names()

baltcity_income <- read_csv("baltcity_income_clean.csv") %>% 
  clean_names()

x2020_dropout_rate <- read_csv("Annual_Dropout_Rate_2020.2.csv") %>% 
  clean_names()

x2020_grade_enrollment <- read_csv("Enrollment_By_Grade_2020.csv") %>% 
  clean_names()

x2020_english_scores <- read_csv("2020_ELPA2_Summary.csv") %>% 
  clean_names()

x2020_diplomas_certificates <- read_csv("2020_Diplomas_and_Certificates.csv") %>% 
  clean_names()

#ct10 is the column with census tracts/geoids
baltimore_census_tracts <- read_csv("Maryland_Census_Data_-_Census_Tracts.csv") %>% 
  clean_names()
```

What areas have the highest median household income?
```{r}
#get the top 10 neighborhoods with the highest median income in 2020
top10_baltcity_income <- baltcity_income %>% 
  select(neighborhood, x2020, census) %>% 
  slice_max(x2020, n=10)

view(top10_baltcity_income)
```
#North Baltimore/Guilford/Homeland has the highest median household income, with $199,531.


What about the lowest?
```{r}
low10_baltcity_income <- baltcity_income %>% 
  select(neighborhood, x2020, census) %>% 
  slice_min(x2020, n=10)

view(low10_baltcity_income)
```
#Upton/Druid Heights is the neighborhood with the lowest median household income, with $13,559. This is about enough to get an older, used car with more than 100,000 miles at CarMax --  https://www.carmax.com/cars?price=9000-14000. 

Let's grab the locations in the city with the top white population as well.
```{r}
#cross reference that with the top 10 places with the most white population

updated_maryland_county_population <- maryland_county_lookup_table %>%
  inner_join(maryland_county_population, by="geoid")

top10_white <- baltcity_race %>% 
  select(census_tract,x2020_white,geoid) %>% 
  slice_max(x2020_white, n=10)

view(top10_white)
```
#Census Tract 2712 has the largest white population. This is the tract that includes Bellona Gittings, The Orchards, Lake Evesham and Rosebank, according to this link: https://data.imap.maryland.gov/datasets/maryland::maryland-census-boundaries-census-tracts-2020/explore?location=39.361786%2C-76.619366%2C15.34.  

Also, let's grab the lowest 10 white population.
```{r}
low10_white <- baltcity_race %>% 
  select(census_tract,x2020_white,geoid) %>% 
  slice_min(x2020_white, n=10)

view(low10_white)
```
#Census Tract 1502 has the lowest white population. This is the Easterwood neighorhood, when using this link to find the census tract area: https://data.imap.maryland.gov/datasets/maryland::maryland-census-boundaries-census-tracts-2020/explore?location=39.361786%2C-76.619366%2C15.34. 

How does this compare to schools with the highest graduation rates?
```{r}
#Filter x2020_diplomas_certificates for Baltimore City, then check on location of the school based on a geoid tracker.
baltcity_grads <- x2020_diplomas_certificates %>% 
  filter(lss_name== "Baltimore City") %>% 
  select(lss_name,school_name, total_hs_completers) %>% 
 slice_max(total_hs_completers, n=10)

view(baltcity_grads)
```
#The Baltimore School for the Arts has the highest count of diplomas/certificates out of schools in Baltimore City. Its census tract is not within the top 10 tracts with the largest white population, when using the name of the school looked up at this link: https://data.imap.maryland.gov/datasets/bbe7d09a81fc40c8a7c9f4c80155842e/explore?location=39.289740%2C-76.639600%2C16.69. 

How are English Language scores in Baltimore City? Is there a specific school that has the lowest scores, and in what area of the city is that located? Do students who speak another language get the help they need for high school graduation? 

```{r}
#find lowest English scores by city
x2020_english_scores$proficient_count<- as.numeric(x2020_english_scores$proficient_count)

x2020_baltcity_english_scores <- x2020_english_scores %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, proficient_count) %>% 
  slice_min(proficient_count, n=10)

view(x2020_baltcity_english_scores)
```
#Abbottston Elementary School has the least number of students proficient in English. Using this link (https://data.imap.maryland.gov/datasets/bbe7d09a81fc40c8a7c9f4c80155842e/explore?location=39.324780%2C-76.600300%2C16.69), the geoid (24510090500) is not either in top10_white or low10_white.

Now to find the schools with the highest proficiency count in Baltimore City.
```{r}
x2020_english_scores$proficient_count<- as.numeric(x2020_english_scores$proficient_count)
  
x2020_baltcity_english_scores <- x2020_english_scores %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, proficient_count) %>% 
 slice_max(proficient_count, n=10)

view(x2020_baltcity_english_scores)
```
#John Ruhrah Elementary/Middle has the highest English scores in Baltimore City. This means that it has the most students who tested to be proficient in their English exams, or ELPAs. The school's geoid (24510260501) based on https://data.imap.maryland.gov/datasets/bbe7d09a81fc40c8a7c9f4c80155842e/explore?location=39.324780%2C-76.600300%2C16.69 also does not correspond with the communities with the most or least white populations.

What schools in Baltimore City have the highest drop out rate in 2020?
```{r}
x2020_dropout_rate$dropout_rate_numer <- as.numeric(x2020_dropout_rate$dropout_rate_numer)
  
baltcity_dropout <- x2020_dropout_rate %>% 
  filter(lss_name=="Baltimore City") %>% 
  select(school_name, dropout_rate_numer) %>% 
  slice_max(dropout_rate_numer, n=10)

view(baltcity_dropout)
```
#Excel Academy has the highest dropout count besides from the count for the entirety of Baltimore City schools. This means that the academy has the highest number of students who have dropped out in the entire Baltimore City school system. 145 students is about the same as a high school graduating class in some smaller, rural areas.

Let's make a table with findings of the Baltimore City dropout count:
```{r}
baltcity_dropout %>% 
  ggplot(aes(x=school_name, y=dropout_rate_numer, weight=dropout_rate_numer)) +
  theme_economist() +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
    labs(
    title="Schools by dropout rate Baltimore City",
    x = "School Name",
    y = "Dropout Count",
    caption = "source: Baltimore City data")
  
```
#Analysis

#Excel Academy, a public high school within Francis M Wood High School, has about double the 
