#lab_08
Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to:
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings
**Task**: Load janitor and the tidyverse and sf

```{r}
library(janitor)
library(tidyverse)
library(sf)
library(leaflet)
```


## Load data

You have been given two spatial dataframes and one regular dataframe to load.  The dataframes are:

* baltimore_homicides.rds | spatial | location and details about each Baltimore homicide between 2007 and 2017, including information about the status (open, closed with arrest, closed without arrest) and the exact location. Details on the data: https://github.com/washingtonpost/data-homicides
* baltimore_tract_shapefiles.rds | spatial | polygons of each census tract in Baltimore
* baltimore_tract_demographics.rds | regular | the total population and white population of each baltimore Census tract in 2017.

**Task** Load the data

```{r}
baltimore_homicides <- read_rds("data/baltimore_homicides.rds") %>% 
  clean_names()

baltimore_demographics <- read_rds("data/baltimore_tract_demographics.rds") %>% 
  clean_names()

baltimore_shapefiles <- read_rds("data/baltimore_tract_shapefiles.rds") %>% 
  clean_names()

```


## Questions

**Q1** Are homicides that occur in majority white Census tracts in Baltimore solved at a higher rate than homicides in non-majority white Baltimore census tracts?

```{r}
#create values for majority white areas
majority_white_balt <- baltimore_demographics %>% 
  mutate(pct_white = (white_pop/total_pop)*100) %>% 
  mutate(neighborhood_dem = case_when(
    pct_white > 50 ~ "majority_white",
    TRUE ~ "not_majority_white"
  ))

view(majority_white_balt)
```

```{r}
#now, let's filter for majority white areas
filtered_majority_white <- majority_white_balt %>% 
  filter(neighborhood_dem == "majority_white")

view(filtered_majority_white)
```

```{r}
#spatial joining 
shootings_by_neighborhood <- baltimore_homicides %>%
  st_join(majority_white_tracts) %>%
  as_tibble()

```

```{r}
#calculate percentage in majority white and non white neighborhoods

case_status_by_neighborhood <- shootings_by_neighborhood %>%
  group_by(neighborhood_type,disposition) %>%
  count() %>%
  pivot_wider(names_from=disposition, values_from=n) %>%
  clean_names() %>%
  mutate(total_shootings = closed_by_arrest+closed_without_arrest+open_no_arrest) %>%
  mutate(unsolved = open_no_arrest + closed_without_arrest) %>%
  mutate(pct_unsolved = unsolved/total_shootings*100)

case_status_by_neighborhood

```
```{r}
#map
ggplot() +
  geom_sf(data=baltimore_shapefiles) +
  geom_sf(data=baltimore_homicides)

```



* Write up your findings in three paragraphs that could be dropped into a news story, with the answer to the question as the lede. 
* Include any other relevant details.
* Output a table with relevant results. 
* Output at least one map visualization pertinent to your finding. 

**A1** 
Homicides in Baltimore City in majority white neighborhoods are solved almost 20% more than homicides in non-white neighborhoods from 2007 to 2017, according to an analysis of the city's data.

Majority white neighborhoods, classified as having more than 50% of a white population, had about 46% of homicide cases unsolved. This is in comparison to non-white neighborhoods, which had about 66% of homicides unsolved. 

Majority white neighborhoods also had the least number of total shootings -- non-white areas had more than 10 times the number of shootings in the same decade, according to data analysis. More than 10 times the number of cases in non-white neighborhoods also went unsolved in comparison to predominantly white neighborhoods, according to the analysis. 
