**Lizzy Alspach**
*JOUR472 Final Project Milestone 2*
*04/25/2024*

This summer, I’ll be interning with The Baltimore Sun and covering Carroll and Howard counties. Because of the large number of book bannings and other factors affecting education, I want to do more data analysis into those fields. I also want to explore how Hispanic/Latino populations have changed in these counties, what with the rising cost of living. 

First, we need to install our packages.
```{r}
library(tidyverse)
library(tidycensus)
library(rvest)
library(janitor)
```


Now, to answer some of my questions about Howard and Carroll county demographics, I need to grab Census data. I'll take it with my API key. 
```{r}
#census key
census_api_key("423292f0a0d4b5f0fe4ce722f283f6af07c21c84", install=TRUE, overwrite=TRUE)

readRenviron("~/.Renviron")
```

```{r}
acs_vars <- load_variables("acs5", year = 2020)
```


Next, I need to specifically grab the sections for Howard and Carroll counties.
```{r}
domestic_consump_prod_carroll <- read.csv("domestic_production_consumption_carroll.csv") %>% 
  clean_names()

view(domestic_consump_prod_carroll)

employment_carroll <- read.csv("industry_employment_carroll.csv") %>% 
  clean_names()

view(employment_carroll)
```

